Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-10-29 summary.Rmd
minor tweaks to writeup

For estimating $\{\lambda^{(kl)}_K$, both methods usethe ground truth community 
labels.

@@

* Number of vertices $n = 256, 512, 1024, 2048, 4096$
* Number of underlying communities $K = 2, 3, 4$
* Mixture parameters $\alpha_k = 1 / K$ for $k = 1, ..., K$, (i.e., each 
community label has an equal probability of being drawn)
* Community labels 
$z_k \stackrel{iid}{\sim} Multinomial(\alpha_1, ..., \alpha_K)$
* Within-group popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

@@

```{r clust_err_k, fig.cap = 'IQR of clustering error using Algorithm 2 (red) compared against modularity maximization (blue) and sparse subspace clustering (green) for sample sizes from 128 to 4096 and number of underlying communities from 2 to 4. Simulations were repeated 50 times for each sample size.', fig.width = 10}
clustering.df <- readr::read_csv('clustering-k.csv')
clustering.df %>%
  dplyr::group_by(K, n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm,
                colour = 'Modularity maximization')) +
  geom_errorbar(aes(x = n, ymin = first.q.mm, ymax = third.q.mm,
                    colour = 'Modularity maximization'), width = .1) +
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') +
  scale_y_log10()
```

@@

```{r clust_err_ct_sim, fig.cap = 'IQR of counts of misclustered vertices using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 128 to 4096. Simulations were repeated 50 times for each sample size.', cache = FALSE, fig.width = 10}
clustering.df <- readr::read_csv('clustering-k.csv')
clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'misclustering count', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm * n,
                colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm * n, ymax = third.q.mm * n,
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') + 
  scale_y_log10()
```

@@

* Number of vertices $n = 256, 512, 1024, 2048, 4096$
* Number of underlying communities $K = 2, 3, 4$
* Mixture parameters $\alpha_k = \frac{k^{-1}}{\sum_{l=1}^K l^{-1}}$ for 
$k = 1, ..., K$
* Community labels 
$z_k \stackrel{iid}{\sim} Multinomial(\alpha_1, ..., \alpha_K)$
* Within-group popularities $\lambda^{(kk)} \stackrel{iid}{\sim} Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \stackrel{iid}{\sim} Beta(1, 2)$ for $k \neq l$

@@

Fig \ref{fig:clust_err_k_imba} and \ref{fig:clust_err_ct_k_imba} show 
similar results as in the balanced communities case, with both Algorithm 2 and 
Sparse Subspace Clustering resulting in no misclustered vertices for a 
sufficiently large sample size. However, Fig \ref{fig:lambda_est_k_imba} 
suggests that while Algorithm 5 retains $\sqrt{n}$ efficiency, the MLE-based 
plug-in estimator fails to do so.

```{r clust_err_k_imba, fig.cap = 'IQR of clustering error using Algorithm 2 (red) compared against modularity maximization (blue) and subspace clustering (green) for sample sizes from 128 to 4096 and number of clusters from 2 to 4. Simulations were repeated 50 times for each sample size.', fig.width = 10}
clustering.df <- readr::read_csv('clustering-k-imbalanced.csv')
clustering.df %>%
  dplyr::group_by(K, n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm, colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm, ymax = third.q.mm, 
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') +
  scale_y_log10()
```

```{r clust_err_ct_sim_imba, fig.cap = 'IQR of counts of misclustered vertices using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 128 to 4096. Simulations were repeated 50 times for each sample size.', cache = FALSE, fig.width = 10}
clustering.df <- readr::read_csv('clustering-k-imbalanced.csv')
clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'misclustering count', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm * n,
                colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm * n, ymax = third.q.mm * n,
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') + 
  scale_y_log10()
```

@@

```{r, cache = FALSE, fig.cap = 'Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by Algorithm 2 (left), sparse subspace clustering (middle), and modularity maximization (right).', fig.height = 2, fig.width = 8, eval = TRUE}
K <- 2
hh.12 <- readRDS('data/village-hh-12.rds')
hh.31 <- readRDS('data/village-hh-31.rds')
hh.46 <- readRDS('data/village-hh-46.rds')

@@

knitr::kable(data.frame(Method = c('Alg. 2', 'SSC', 
                                   'Mod. Max.'),
                        Error = round(1 - c(hh.acc.12, hh.acc.ssc.12, hh.acc.mm.12), 3)),
             caption = paste('Community detection error rates for identifying',
                             'household religion.'))
```

>>>>

For estimating $\{\lambda^{(kl)}\}_K$, both methods usethe ground truth community 
labels.

@@

* Number of vertices $n = 128, 256, 512, 1024, 2048, 4096$
* Number of underlying communities $K = 2, 3, 4$
* Mixture parameters $\alpha_k = 1 / K$ for $k = 1, ..., K$, (i.e., each 
community label has an equal probability of being drawn)
* Community labels 
$z_k \iid Multinomial(\alpha_1, ..., \alpha_K)$
* Within-group popularities $\lambda^{(kk)} \iid Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \iid Beta(1, 2)$ for $k \neq l$

@@

```{r clust_err_k, fig.cap = 'IQR of clustering error using Algorithm 2 (red) compared against modularity maximization (blue) and sparse subspace clustering (green) for sample sizes from 128 to 4096 and number of underlying communities from 2 to 4. Simulations were repeated 50 times for each sample size.', fig.width = 10}
clustering.df <- readr::read_csv('clustering-k.csv')
clustering.df %>%
  dplyr::group_by(K, n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error rate', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm,
                colour = 'Modularity maximization')) +
  geom_errorbar(aes(x = n, ymin = first.q.mm, ymax = third.q.mm,
                    colour = 'Modularity maximization'), width = .1) +
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') +
  scale_y_log10()
```

@@

```{r clust_err_ct_sim, fig.cap = 'IQR of counts of misclustered vertices using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 128 to 4096. Simulations were repeated 50 times for each sample size.', cache = FALSE, fig.width = 10}
clustering.df <- readr::read_csv('clustering-k.csv')
clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error count', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm * n,
                colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm * n, ymax = third.q.mm * n,
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') + 
  scale_y_log10()
```

@@

* Number of vertices $n = 128, 256, 512, 1024, 2048, 4096$
* Number of underlying communities $K = 2, 3, 4$
* Mixture parameters $\alpha_k = \frac{k^{-1}}{\sum_{l=1}^K l^{-1}}$ for 
$k = 1, ..., K$
* Community labels 
$z_k \iid Multinomial(\alpha_1, ..., \alpha_K)$
* Within-group popularities $\lambda^{(kk)} \iid Beta(2, 1)$
* Between-group popularities 
$\lambda^{(kl)} \iid Beta(1, 2)$ for $k \neq l$

@@

Fig \ref{fig:clust_err_k_imba} and \ref{fig:clust_err_ct_sim_imba} show 
similar results as in the balanced communities case, with both Algorithm 2 and 
Sparse Subspace Clustering resulting in no misclustered vertices for a 
sufficiently large sample size. However, Fig \ref{fig:lambda_est_k_imba} 
suggests that while Algorithm 5 retains $\sqrt{n}$ efficiency, the MLE-based 
plug-in estimator fails to do so.

```{r clust_err_k_imba, fig.cap = 'IQR of clustering error using Algorithm 2 (red) compared against modularity maximization (blue) and subspace clustering (green) for sample sizes from 128 to 4096 and number of clusters from 2 to 4. Simulations were repeated 50 times for each sample size.', fig.width = 10}
clustering.df <- readr::read_csv('clustering-k-imbalanced.csv')
clustering.df %>%
  dplyr::group_by(K, n) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error rate', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm, colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm, ymax = third.q.mm, 
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') +
  scale_y_log10()
```

```{r clust_err_ct_sim_imba, fig.cap = 'IQR of counts of misclustered vertices using Algorithm 2 (red) compared against subspace clustering (blue) for sample sizes from 128 to 4096. Simulations were repeated 50 times for each sample size.', cache = FALSE, fig.width = 10}
clustering.df <- readr::read_csv('clustering-k-imbalanced.csv')
clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75),
                   med.err.ssc = median(error.ssc),
                   first.q.ssc = quantile(error.ssc, .25),
                   third.q.ssc = quantile(error.ssc, .75),
                   med.err.mm = median(error.mm),
                   first.q.mm = quantile(error.mm, .25),
                   third.q.mm = quantile(error.mm, .75)) %>%
  dplyr::ungroup() %>% 
  ggplot() +
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
  labs(y = 'community detection error count', 
       colour = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Algorithm 2')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Algorithm 2'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'SSC')) + 
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'SSC'), width = .1) + 
  geom_line(aes(x = n, y = med.err.mm * n,
                colour = 'Modularity maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.mm * n, ymax = third.q.mm * n,
                    colour = 'Modularity maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both') + 
  scale_y_log10()
```

@@

```{r, cache = FALSE, fig.cap = 'Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by Algorithm 2 (left). The villages studied here are, from left to right, 12, 31, and 46.', fig.height = 2, fig.width = 8, eval = TRUE}
K <- 2
hh.12 <- readRDS('data/village-hh-12.rds')
hh.31 <- readRDS('data/village-hh-31.rds')
hh.46 <- readRDS('data/village-hh-46.rds')

@@

hh.acc.31 <- mean(zhat.hh.31 == z.hh.31)
hh.acc.31 <- ifelse(hh.acc.31 < .5, 1 - hh.acc.31, hh.acc.31)
hh.acc.ssc.31 <- mean(zhat.hh.ssc.31 == z.hh.31)
hh.acc.ssc.31 <- ifelse(hh.acc.ssc.31 < .5, 1 - hh.acc.ssc.31, hh.acc.ssc.31)
hh.acc.mm.31 <- mean(zhat.hh.mm.31 == z.hh.31)
hh.acc.mm.31 <- ifelse(hh.acc.mm.31 < .5, 1 - hh.acc.mm.31, hh.acc.mm.31)

hh.acc.46 <- mean(zhat.hh.46 == z.hh.46)
hh.acc.46 <- ifelse(hh.acc.46 < .5, 1 - hh.acc.46, hh.acc.46)
hh.acc.ssc.46 <- mean(zhat.hh.ssc.46 == z.hh.46)
hh.acc.ssc.46 <- ifelse(hh.acc.ssc.46 < .5, 1 - hh.acc.ssc.46, hh.acc.ssc.46)
hh.acc.mm.46 <- mean(zhat.hh.mm.46 == z.hh.46)
hh.acc.mm.46 <- ifelse(hh.acc.mm.46 < .5, 1 - hh.acc.mm.46, hh.acc.mm.46)

knitr::kable(dplyr::tibble(
  Method = c('Alg. 2', 'SSC', 'Mod. Max.'),
  `Error (Village 12)` = round(1 - c(hh.acc.12, hh.acc.ssc.12, hh.acc.mm.12), 3),
  `Error (Village 31)` = round(1 - c(hh.acc.31, hh.acc.ssc.31, hh.acc.mm.31), 3),
  `Error (Village 46)` = round(1 - c(hh.acc.46, hh.acc.ssc.46, hh.acc.mm.46), 3)),
  caption = paste('Community detection error rates for identifying',
                  'household religion.'))
```
